A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-31 latex/main.tex
Updates from Overleaf

We close by emphasizing several current limitations
as well as potential directions to improve sc2ts,
which we hope will also inspire future research
to develop ARG methods for other pathogens.

Deletions which are reverted can be strong signals of recombination,
because bases which are lost cannot be easily regained,
except through recombination involving
a lineage that has bases at the deleted sites.
We currently treat multi-base deletions as missing data.
More sophisticated treatments of such indels in the HMM
may help recover additional signals
(and avoid false signals) of recombination.

The current HMM assumes a simplistic mutation model
(where mutation to any other allele is equally probable) and
implicitly assumes a uniform mutation rate along the SARS-CoV-2 genome.
However, genome analyses have revealed
transition-to-transversion rate bias \cite{Bloom2023,Simmonds2024} and
sequence context-dependent variation
in the mutation rate in the SARS-CoV-2 genome \cite{Haddox2025}.
A more realistic mutation model may help improve
inference of local genetic relationships in the ARG, and
accounting for mutation rate heterogeneity
during sample matching or in the HMM cost
(e.g., down-weighing sites in highly mutable regions)
may help reduce false positive recombinants.

Finally, future developments may account for geography in the HMM,
favoring attachment of samples to geographically close lineages.

@@

It is challenging to accurately reconstruct global events
when there are regions with poor sampling and
jurisdictions with long delays
before sequences become publicly available (REF).
For example, revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating
with samples that became available later)
might improve the accuracy of the genealogy.

Several methods are developed and used
to exclude probable false positive recombination events \textit{post hoc},
including filtering by supporting loci per parent and
splitting of suspiciously long branches.
These methods could instead be run immediately
after sample matching during ARG inference,
obviating the need for post-processing of the ARG.

>>>>

We close by emphasizing several current limitations
as well as potential directions to improve sc2ts,
which we hope will also inspire future research
to develop ARG methods for SARS-CoV-2 and other pathogens.

Reversions of deletions can indicate recombination,
because bases that are lost are unlikely to be regained,
except via recombination with a lineage that has those bases.
We presently treat multi-base deletions as missing data.
Explicitly modeling deletions in the HMM
could help recover additional signals of recombination
and reduce false positives.

The current HMM assumes a simplistic mutation model
(where mutation to any other allele is equally probable) and
implicitly assumes a uniform mutation rate along the SARS-CoV-2 genome.
However, genomic analyses have revealed
transition-to-transversion rate bias \cite{Bloom2023,Simmonds2024} and
sequence context-dependent mutation rate variation
across the SARS-CoV-2 genome \cite{Haddox2025}.
Using a more realistic mutation model and
accounting for mutation rate heterogeneity
could improve
inference of local genetic relationships in the ARG and
reduce spurious recombinants.

Future developments may account for geography in the HMM,
favoring attachment of samples to geographically close lineages.

@@

We have developed and used filters
to detect time traveling samples in the Viridian dataset and
to exclude them from the ARG,
yielding a set of samples with reliable collection dates.
Future studies could leverage the ARG
to flag wrongly dated samples or
to infer the dates of samples
with missing or incomplete collection dates in the metadata.

It is challenging to accurately reconstruct global events
when there are regions with poor sampling and
jurisdictions with long delays
before sequences become publicly
available \cite{Lancet2021,Kalia2021,Chen2022}.
For example, revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating
with samples that became available later)
might improve the accuracy of the genealogy.

We have developed and used several methods
to exclude spurious recombination events \textit{post hoc},
including filtering by supporting loci per parent and
splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
These methods could instead be run immediately
after sample matching during ARG inference,
reducing the number of ARG post-processing steps.
